# Introduction to R and RStudio
### Jonathan Godfrey and Jonathan Marshall

##  Getting started with R

The standard method for running R is the Graphic User Interface (GUI). The standard installation on a  64-bit Windows machine puts two shortcuts on the desktop; one is for 32-bit operation. It won't matter which we use in this first session, or for most of the work you will do for some time!

If you haven't done so before, investigate the menus. Pretty soon you will realise that R doesn't give you menus for doing any analyses. The menus are useful for managing your session, your additional packages (more on them later), and accessing the help and other documentation.

## Methods for running R 

Ultimately we will make use of the most commonly used integrated development environment (IDE) in use today (RStudio) but there are so many ways of running R jobs. After this introductory workshop you will need to decide which methods suit you and your work the best; you will also find that this decision may need to be revisited later.


### The R GUI (Console)


Open the R GUI now. We'll do some very basic tasks to investigate the different ways of working.

#### Using the command prompt (laborious typing?)


The opening screen has a welcome message and then the prompt where we can type commands. I always start with the things we know, like `1+1` whenever I use any new software. Let's do something a little more interesting. Try:

```{r RandomSample}
sample(100,10)
MySample = sample(100,10)
MySample
```

The `sample` function grabs a random sample that is as long as the second argument, from the first argument; in this case the first 100 integers. (Try `sample(letters, 10)` to confirm this.)  Over the years, I've needed to remind people that the sample gets re-drawn each time the command is called, but that we can keep a permanent copy if we **assign** the result. 

For the purposes of the illustration, we do want to do something with our sample. For example we might want to sort the results or calculate their mean. Let's do so but use two useful features - tab completion and arrowing back through our history. When you type the first command in the next block, do not type `MySample` out in full. Type one or two characters and press the `tab` key. Instead of typing the second command out in full, press the up and down arrow keys to find a command that is suitable for editing. Both of these ideas are good for avoiding typos.

```{r PlaySample}
sort(MySample)
mean(MySample)
```

The command prompt is a great way of experimenting, but it really is not a great place to be efficient. If you do find your experiments getting out of hand, save your work using the command `savehistory("SomeFilename.R")` and then ask where you just saved that file using `getwd()` - we'll want to browse for that file very soon so remember the answer please.

After saving your history, close the R session. Please say `yes` to the save message.

Browse to the working directory. Look for three files: `SomeFilename.R` (an R script), `.RData` (the data saved from the work in your session), and `.Rhistory` (the full history of commands used in that workspace).

Clicking on `.Rhistory` should have no impact; clicking on `SomeFilename.R` might do something strange, but hopefully won't; and clicking on `.RData` will open the R GUI again. You can move `.RData` and `.Rhistory` to a different folder so that work from this session can be kept separate from other work. Use of different folders for different projects is **very** sensible. It also means you could take those two files to another computer.

Make a new folder please and move the three files to it now. Then re-open the R GUI by clicking on the `.RData` file. You could confirm the new working directory using `getwd()` if you like. (Just press the up arrow to get back through the history.)

#### The R script window

The R GUI has its own basic text editing window for script files. There is a menu item to open the window with a blank space, as well as the standard item to open the script file.


Please grab the script file we created earlier. Do some minor editing, and save it again.

You can run the entire script, or various parts of it using menu items in the `Edit` menu.

Now close the script window.

This method of working is quite common, but has perhaps been surpassed by better options. 

#### Running an R script without opening it

The R script you just closed can be pulled back into any R session using the `source()` command. 
In full, that would be `source("SomeFilename.R")`. (You should be getting into the habit of tab completion.) 

### The R Terminal

This is **very** old school! This method of working is used by a minority of people. Everything a user could do in the terminal is achievable in the GUI console. You probably won't make use of this method of working if you are a windows user, but many Linux users are using a terminal so it is worth knowing about.

Windows users are not given immediate access to the terminal version but it can be accessed just as easily as the GUI once we make a new shortcut for it. 

- Copy and paste the R GUI shortcut on your desktop. 
- Edit the target this new shortcut points towards. It currently says `...\rgui.exe"` to open the GUI console. Delete the `gui` and insert `term` in its place so that it reads `...rterm.exe"` instead.
- It's probably a good idea to rename this new shortcut; use the f2 key.
- Open the terminal using this new shortcut

You should see the same welcome screen as you saw in the GUI console except that there are no menus and no colours. There will be no script editing window, but access to help facilities and the like is the same as for the R GUI.


You are almost certainly in the same working directory as you were for the exercises done previously. (You could check.)

Use of the terminal in conjunction with a text editor for creating scripts is perfectly acceptable. (It's what I do - JG). It works because all functionality offered in the R GUI can be achieved using commands alone.

### Using the command line to run jobs in batch mode


Any script file can be pushed into an R session without opening the application. The method for Linux is much better than for Windows, but it's still a useful process under Windows.

Why use batch mode? A  script file that is run using batch mode can be set up as a scheduled  task. This could prove very useful if a task needs to be repeated periodically to incorporate the latest data for example.

#### Batch jobs under Windows

If we know the full path to the main executable files for R, we can create a command line that accepts the name of the R script as an argument. It will look something like:

`C:\PROGRA~1\R\R-32~1.1\bin\i386\R.exe CMD BATCH --vanilla --quiet MyScript.R`

The `vanilla` switch says to process `MyScript.R` in a brand new R session, and `quiet` means the returned file of output will not include the welcome messages seen at the start of any R session.

The resulting output file will be a plain text file called `MyScript.Rout` so will not include any graphs, which would need to be saved explicitly within the script.

The above command line can be put into a batch file which is just a text file with the extension `.bat` that windows knows how to process. This batch file is automatically executable under Windows so it can be clicked on within Windows Explorer.

 


#### Batch Jobs under Linux

The process is even simpler for Linux users who just add the following line to the very start of their R script files.

`#!/rscript`

and then make that script file executable using standard bash commands. This process is being used to run the main script file for RenalQ every 15 minutes. The above line is called a **shabang** line.


###  RStudio


This is the most common working environment outside base R. We will go into more detail later in this session, but we need to illustrate the alternatives in order to understand the benefits of working with RStudio.

### Other front-ends

Revolution Analytics (now owned by Microsoft) has a product called Revolution R Open (RRO). On the face of it, RRO is just the R GUI (64-bit) with some hidden benefits. The marketing suggests improved methods for handling some calculations will be employed. It is questionable how useful we will find this product.

Open Analytics has an IDE called Architect. It is based on the Eclipse IDE. If you are experienced with Eclipse environments, then feel free to investigate this product. Be warned that installation on my laptop  hasn't been plain sailing (JG) so it is mentioned for information only at this time.

The R commander and Deducer front ends add menus to the standard R GUI to provide a range of statistical analyses in dialogue box form. They can be useful for the novice user, but they do have some problems/limitations to note. First, they do not leave the user with an R script file that can be used by any other user, that is, they do not always generate standard R code for every task. Second, they are not comprehensive and so cannot be relied upon for full integration with other actions and methods of working. It is hoped that these workshops push past the *novice* level and that their usefulness is therefore less relevant to your team.
  
## Some basic things we ought to know

This is not an exhaustive list. If you ask questions there will be answers.

### Help pages and documentation

There are two ways to get help from R. Either search for help on a specific function you know the name of (or are willing to guess), and searching for a word that appears on a help page. These ideas use question marks (1 and 2 respectively). The main problems with this help system is the fact that you must know the function's name, or be good at selecting a suitable term to search for

The help pages are fairly good for illustrating the usage of a command. The examples are sometimes concise (a.k.a. terse)  and informative, but there are plenty of examples that do considerably more than offer simple demonstrations.

There are plenty of add-on packages that you might use. These often include demonstrations, but of late the culture is to create **vignettes** to illustrate the use of the additional package.

A major advantage of the help page examples and vignettes is that the R code in them must be syntactically correct. (There are situations where the development process hasn't used the testing features. Look out for parts of the examples on help pages where it says `dontrun` or `donttest`).

#### R bloggers

This is  an aggregator  service that monitors over 400 blogs that pertain to R. It supplies me with a daily distraction by email, but is also a good way for me to keep abreast of developments in the way people are using R, demonstrating packages and analysis ideas, and promoting activities of interest to the R community. The main site is searchable and is therefore a good source of information.

#### Stack overflow

Most of the time that I resort to google for the answer to a question that relates to R, I find the results point to stack overflow more than anywhere else. If you do post to stack overflow, the readers expect you to have an example that they can reproduce for testing purposes. Failure to do so can lead to some grumpiness, and a lack of responses.

### Packages (installing and loading)

Add-on packages are really important. The first one we will look at helps keep your Windows installation of R up-to-date.

We will download the `installr` package now. Downloading is the first part of installing a package. This just puts it within the R framework for later use. It does not mean you have immediate access to the package functionality.

There is a chance that a proxy server will interfere with this process. As an exercise in paranoia, I have tended to automatically issue the command:

```{r setInternet2}
setInternet2()
```

because students don't always know what a proxy server is. This command has no impact if no proxy server is in place.

Then we need to look for a CRAN mirror. This is one of the servers around the world that holds the R packages. There is a menu item, but I use the command:

```{r chooseCRANmirror}
chooseCRANmirror(ind=1)
```

which selects the first item off that list. It is the cloud-based server operated by RStudio. This is a redirection service and helps find the closest server to us.

Then, we would install the desired package using the menu item or:

`install.packages("installr")`
 
I prefer to type the package name as looking down the list of over 6000 packages takes longer than typing the command.

To get this package loaded for use, we need to use either the `require` or `library` commands:

```{r loadPackage}
library(installr)
```


The command that can be run occasionally is 
`updateR()`

It will check that the current version of R is the most recent, and update accordingly. This includes managing the removal of the old version, updating all packages, and moving the packages you used in the past to the right location for the new version of R. Each step is controlled by the user, although the particular choices can be added to the above command to more fully automate the process.

## Setting up R for personal/project work


I recommend that every new piece of work be done in a brand new workspace (folder). This means that changing to that directory within R and saving your workspace leaves you with compartmentalised workspaces. You could then go back to exactly where you left off on any particular work. It means you can share the workspace with confidence that it contains only what the other person  needs or you want to share.

You can customize R so that it has some actions taken as you open the application no matter which folder you are working in. I recommend however that you do this within each folder or workspace rather than across all workspaces. The simplest way is to construct a `.Rprofile` file that lists the R commands you want to be run as R is opened in this workspace.

For example, I wanted the `knitr` package to be ready for use in the workspace for this set of workshops. My `.Rprofile` has the single command `library(knitr)` in it so I need not issue this command each time I come back to this workspace.

To do this, we need to write the single line to the text file called `.Rprofile` using:
 
```{r makeRprofile}
cat("library(knitr)\n", file=".Rprofile")
```

### Important resources

There are hundreds of books (free and costly) that use R. Purchase with caution.
Some of these books are written to support particular add-on packages, while others use add-on packages to help package the work for the text together.
Sorry, there is no useful advice to share on how useful a  book will be for you; we're just too different and we can't know everything about every book. The best you'll get is something like "you might look at..."!

So, you might look at the set of notes I have for doing tasks. It is called "Let's Use R Now", or "LURN" for short. It was written to help introductory students by having chapter and section headings that are task oriented and an index for searching out functions and ideas. When I started this project there were very few free books that were as easily searched.

Visit the [LURN front page](http://r-resources.massey.ac.nz/lurn/front.html) now to see what tasks are already incorporated. We'll probably add to the notes given there for some of the exercises to be done in future workshops.

#### Importing data

There are so many ways data are stored, and some are specific to particular software used by other people. R can import practically any data set from a wide variety of sources (csv file, URL, delimited files, from packages, and xls). These are covered in a chapter of LURN.


## Useful links for setting up a new installation

R is  available from [CRAN - the Comprehensive R Archive Network.](http://www.r-project.org) 
The following is a  direct link for Windows users  to the [latest installer for R under Windows XP, Vista, 7, and 8.](http://cran.rstudio.com/bin/windows/base/release.htm)

RStudio is available from the [RStudio download page.](http://www.RStudio.com/products/rstudio/download)

pandoc is available from the  [pandoc download page.](https://github.com/jgm/pandoc/releases)

An installation of miktex is recommended for anyone wishing to incorporate R output and graphs in a pdf file. It is available from the [miktex home page](http://www.miktex.org)

## Using RStudio

RStudio is the most popular integrated development environment (IDE) for working with R. An integrated development environment generally combines a good text editor that knows quite a bit about the language being written (in this case R) with tools that make managing the development process (writing code across multiple files) a little easier. In the case of RStudio we have

### The console

By default, in RStudio you'll see the Console on the left of the screen. This is pretty much identical to the RGui console windows, though the colours may differ a little bit. You can type stuff in here, or copy and paste stuff into here the same way as you can with plain R terminal or R Gui to execute R code directly.

### Environment and History tabs

The Environment and History tabs (as well as more that we'll get to later) can be found in the top right in RStudio.

#### Environment

For now, you can think of the Environment tab as holding your workspace (the 'Global Environment'). It'll have any data you have loaded in the form of data-frames, plus any additional non-data-frame variables listed in the Values section, and any functions that R knows about (that you've written) in the Functions section. In addition, you can take a look at what is inside any packages you have loaded by clicking on the drop down above. There's also shortcuts toolbar icons for loading and saving the workspace, importing data, and clearing the workspace.

If you re-execute the code you wrote earlier, you'll have your `MySample` object showing up, which you can see is stored as an integer with indicies 1 through 10, with the values alongside.

For data frames, or for things that are lists in R, there's a little more functionality available. Let's create one so we can take a look. A data frame is just a bunch of columns the same length. i.e. just like you'd normally have in a set of data.

```{r}
MyDataFrame <- data.frame(num = 1:10, alpha = letters[1:10])
MyDataFrame
```

If you execute the above, you'll find you'll have `MyDataFrame` in the Data section. You can click on the little arrow next to it on the left to take a look at it's structure: This is equivalent to running `str(MyDataFrame)` from the console:

```{r}
str(MyDataFrame)
```

You can double click on your data frame to get it to show you the data in a spread-sheet like format. With newer versions of RStudio you should be able to re-order or filter the columns in this view which is quite handy just for looking through data.

#### History

The history window shows you your history. It's basically just showing your `.Rhistory` file, but adds a couple of small niceties. You can select lines (or blocks) of code you've previously written and copy them into the console with the 'To Console' button to re-run them which can be handy. Usually more handy is you can copy those lines into an R script with the 'To Source' button. We'll get to R Scripts when we discuss the editor below.

### Files, plots, packages, help

In the bottom right is the Files, Plots, Packages and Help tabs (as well as others we'll get to later!)

#### Files

The Files tab is similar to 'Windows Explorer'. It allows you to navigate the filesystem, open scripts or markdown documents, create folders, or delete and rename files. Under the More toolbar button there's Copy and Move, as well as the ability to set the working directory to a given folder.

#### Plots

The Plots tab will hold any plots you create. You'll be able to go back and forward through your plot history to look at previous ones (by default R will overwrite the current plot when you do a new one in RGui, though ofcourse this can be controlled) as well as export them to an image or PDF file. There's even a 'Publish' button which will push your plot up to RPubs, a (free) public website hosted by RStudio so you can share your plot with the world should you wish.

#### Packages

The packages tab encapsulates the R package manager. Here you can see all the packages you have available which will include system pacakges as well as packages you've installed. Those that are loaded in the current session will have a tick on the left. To load them, just check the box, and RStudio will execute `library(package_name)` for you. To unload, uncheck the box. You can install and update packages with the click of a button here as well. The Update button will present you with all the packages that have an update available, and you can see what has changed and update each of them as you like.

You can also click on the package name to access the help for that package. You might want to try that on a few of the packages to see what sort of help at a package level R provides - that takes you to the next section...

#### Help

The R help system is provided under the Help tab. This allows you to jump through the help (forward/backward) similar to a web-browser, as well as searching the help in general (top right search bar) or within the current topic ('Find in Topic').

### Text editor

What really makes RStudio shine is it's nice integrated text editor. If you don't already have an R Script open, you can open the one you created earlier (`File->Open`) or create a new one (`File->New...->R Script`).

The text editor including in RStudio understands R code. It offers:
- Auto-completion of variables, package and function names, as well as offering hints as to what function parameters can take.
- Auto-indenting of conditional blocks of code (such as inside loops or conditional statements).
- Adding of template `roxygen` comment blocks to assist in documentation.
- Automatic detection of where code will fail, due to incomplete parentheses, commas or brackets.
- Syntax highlighting of R code so that strings, brackets etc. stand out from other code.
- It understands other code formats as well, such as C/C++ and R Markdown.
- It has a bunch of helpers built in, such as taking a set of R commands and turning them into a function, commenting or indenting large chunks of code, re-formatting code into 'tidier" forms.
- It also integrates with the R debugging tools to allow you to easily set breakpoints in code and step through code with the R debugger.
- Lastly, it has some handy toolbar buttons or keyboard shortcuts for running blocks of code, for sourcing R scripts, compiling R Markdown, or creating R notebooks, which translate an R script into an R markdown document and then compiles it.

### RStudio projects

Just like with plain R, it pays to organise your work into folders. RStudio calls these 'Projects' - they're just a folder that can help you organise R scripts or data files and which is used to save your history (`.Rhistory`) and workspace (`.Rdata`) files. In addition, RStudio adds a file named `dir_name.Rproj` which holds extra information that RStudio knows about your project. This can include preferences for the text editor (e.g. the spacing to use for indents, or the character-encoding to use for text containing characters such as umlauts) or other parts of RStudio.

You might like to associate a new RStudio project with the existing folder you've been working in, by going to `File->New Project...` and selecting `Existing Directory`. It will add the `.Rproj` file. You can then exit RStudio and get back into it where you were by clicking on the `.Rproj` file. This will re-load your history and workspace, and set the working directory to that folder.

### Other neat stuff

A lot of the new R development is happening with folk that either work for RStudio directly, or use it as a development environment. Thus, RStudio has quite a few nice-to-haves when writing R code. All of these are available in plain R, so can all be invoked in the same ways above (and there are often situations where this is more efficient!) but for interactive work with R code, the shortcuts provided can be quite nice.  A few examples are listed below.

#### Version control integration

RStudio knows when a folder is under version control using Git, and if so adds a Git tab in the top right, allowing you to see code differences compared to previous versions and add files or commit new versions to a repository. It also allows you to work with remote repositories, such as [Github](http://github.com).

#### Integration with shiny

Shiny is a package that helps produce javascript-based webapps that communicate with an R back-end on a webserver to produce interactive webpages. RStudio can recognise when a shiny app is being developed and adds toolbar buttons to allow you to run the shiny application directly.

#### Package development tools

If you want to develop an R package, RStudio can help to simplify the process by utilising the `devtools` package. It adds a 'Build' tab in the top right that allows you to compile documentation, run tests (using the `testthat` package), build the package, run the CRAN build process (so you can fix warnings prior to submission to CRAN) and so on.
