---
title: "How to weigh a donkey in Morocco"
author: "Joanne Blogs"
output:
  html_document: default
---

## Introduction

In this notebook we'll look at data collected on 385 donkeys used to transport goods to market in Morocco. It is a sample of the data collected for *Estimation of the liveweight and body condition of working donkeys in Morocco* by R.A. Pearson and M. Ouassat, published in the Veterinary Record, 1996. The following variables are available.

Variable   Description
--------   -----------
Sex        Sex (male/female)
Age        Age (years)
Bodywt     Weight (kg)
Heartgirth Girth at the heart (cm)
Umbgirth   Girth at the umbilicus (cm)
Length     Length from elbow to buttocks (cm)
Height     Height to the withers (cm)

We start by reading the data in from the web and looking at the first few rows:

```{r}
library(ggplot2)
donkey = read.csv("http://www.massey.ac.nz/~jcmarsha/227215/data/donkey.csv")
head(donkey)
```

We notice that the donkey data has a bunch of variables. Our interest is predicting body weight (`Bodywt`) using the other, more easily measureable things. Use of the `summary` command will give us some idea of how each of the columns are distributed:

```{r}
summary(donkey)
```

From this we see that the Bodyweight is somewhere between 52 and 222 kg, but that doesn't tell us much if we want to predict the body weight for a particular donkey. For that, we'll need to take advantage of the relationship between body weight and the other measures.

### Relationship between body weight and other measures

Let's see if there is a relationship between body weight and some of the other measurements. e.g. to assess body weight versus height we can use:

```{r}
ggplot(donkey, aes(x=Height, y=Bodywt)) + geom_point()
```

It's clear there is an increasing relationship between Bodyweight and Height as might be expected. This suggests we could use a particular donkey's height to get a better idea of its body weight.
