---
title: "227.215 Biostats Lab 4"
author: "Jonathan Marshall"
output:
  pdf_document: null
  html_document: default
graphics: yes
---

## Introduction ##

This lab is a short one - we'll look at a set of data on plate meter measurements of grass by your peers a few years back, and how we can treat the lack of independence of residuals.

**After** you're done, feel free to work on your assignment.

## Exercise 1: Create a new R script

In this exercise we'll look at some grass measurement data collected by Animal Production students in 2013 using grass plate meters.

1. Create a new R script file, and copy and paste the metadata block from your lab3.R script into it, changing the title.

2. Add the following lines to read in the grass measurement data and take a look at the summary.

    ```{r, eval=FALSE}
    #' Read in data on grass measurements
    grass <- read.csv("http://www.massey.ac.nz/~jcmarsha/227215/data/grass.csv")
    summary(grass)
    ```

3. Add some code and comments to your R script to produce some plots of `metread` versus the covariates `farm`, `paddock`, and `person`.

4. Is there any graphical evidence that the plate meter readings changes between farms, paddocks or the person using the meter? Add some comments to your R script.

5. Build a linear model to assess whether the plate meter readings differ by farm. What is your conclusion? Add some comments to your R script.

6. A number of different students did the measurements. Add some comments to your R script about why this will likely mean the residuals are not independent.

7. One way to fix the lack of independence is to account for the `person` effect in the model. Build a linear model incorporating `person` as well as `farm`. What is your conclusion now about whether plate meter readings differ by farm?

## Exercise 2: Incorporating a random effect for person

The problem with our linear model from part 7 is that by including the person in the model, we can no longer predict what the average measurement will be on those farms should **other** people take them. Our model is restricted to just the people that did the initial measures.

Ideally, we'd want to account for the variation between people while not having to include it in the model for prediction.

We can do this by treating the person variable as **random**. The basic idea is that we have multiple residuals for each person, so the average of the residuals for a particular person could be treated as that persons effect on the measurement.  We basically rewrite the residuals as

$$
\epsilon_{ij} = \mathsf{person}_i + \eta_{ij}
$$

where $\epsilon_{ij}$ is the $j$-th residual of person $i$ from the linear model with just farm in it, $\mathsf{person}_i$ is the average residual for person $i$, and $\eta_{ij}$ is the left-over $j$-th after subtracting out the person level average. That way these left-overs are hopefully independent, as the person effect has been taken out.

We do this in such a way that the mean of the person effects is 0. Thus, to predict for a new person, we just set the person effect to 0 and incorporate the variation in the person effects into prediction intervals.

To do this in R, we use the `nlme` library, which stands for "non-linear mixed effects". The mixed is because we're mixing normal linear model coefficients (fixed effects) with random effects.

1. Start by loading the library in and fitting a model

    ```{r, eval=FALSE}
    #' Load in the nlme library
    library(nlme)
    #' Fit a model incorporating a random effect for person
    mixed <- lme(metread ~ farm, random=~1|person, data=grass)
    ```

    The `random` bit here can be read as specifying separate mean (intercept) for each person.

2. Do an anova and summary table for the mixed model. What is your conclusion?

3. Notice that the summary table gives you some information about your random effects. The `Intercept` value is the standard deviation of the per-person random effect (i.e. between-person variability), while the `Residual` value is the standard deviation of the left-over residuals after accounting for people (i.e. the within-person variability). You should notice that there is more within-person variation than there is between-person variation.

4. Summarise your results in a few comments in your R script file.

5. All the best for the rest of the course! I'd love to get feedback from you on the biostats portion of the course - feel free to tell me what you think via email or an anonymous note under my door! It'd be great to get a representative sample of responses, so it's important that I get the *meh* responses as well as the *awesome* and *you suck* responses! :)