---
title: "Linear models 3"
author: "Jonathan Marshall"
date: ""
output: 
  ioslides_presentation: 
    highlight: tango
    widescreen: true
    css: jm.css
---

```{r setup, echo=FALSE}
library(knitr)
opts_chunk$set(dev.args=list(bg='transparent'), comment='')
col_points <- "#7f577492"
col_dark   <- "#5f4354"
```
## Learning outcomes

- Review interpretation of `summary` and `anova` tables.

- Interactions.

# Review

## Summary table

- Each row is **adjusted** for the other rows.

- The coefficients are thus the effect after adjusting for the others.

- The P-values are whether that effect is significant, after adjusting for the others.

## Summary table for factors

- Each P-value is testing whether that coefficient is zero or not.

- Factors with more than 2 levels get **multiple rows**.

- We want to know if the factor is important, which means testing that multiple coefficients
are zero at the same time.

- We must use an F-test for this, i.e. the `anova` table.

## ANOVA table

- Each row represents a covariate, regardless of whether it's numeric or a factor.

- Can thus be used to test significance of factors directly.

- Order can be important. Each row is adjusted for the ones above it (but not below it).

- Usually order isn't an issue, but if it is, it tells you something.

# Example: Petrels

## Example: Petrels

```{r, echo=FALSE, fig.width=9.5}
petrels <- read.csv("http://www.massey.ac.nz/~jcmarsha/227212/data/petrels.csv")
petrels$Area <- as.factor(petrels$Area)
petrels$Sex <- as.factor(petrels$Sex)
levels(petrels$Sex) <- c("Male", "Female")
par(mar=c(5,5,1,0), cex.axis=1.5, cex.lab=1.5)
plot(R.Wing.Lth ~ Area, data=petrels, ylab="Right wing length (cm)")
```

## Example: Petrels

```{r, echo=FALSE, fig.width=7, fig.align='center'}
par(mar=c(5,5,1,0), cex.axis=1.5, cex.lab=1.5)
plot(R.Wing.Lth ~ Sex, data=petrels, ylab="Right wing length (cm)", xlab="", las=1, horizontal=TRUE)
```

## Petrels: ANOVA table

```{r}
mod = lm(R.Wing.Lth ~ Sex + Area, data=petrels)
anova(mod)
```

## Petrels: Summary table {.smaller}

```{r, echo=FALSE}
summary(mod)
co <- round(coef(mod),2)
```

## Petrels: Conclusion

- From ANOVA table, both `Area` and `Sex` are important to the right wing length.

- From summary table, we see that males have `r -co["SexFemale"]`cm larger right wing length than females **after accounting for differences between areas**.

```{r, echo=FALSE}
mf = tapply(petrels$R.Wing.Lth, petrels$Sex, mean, na.rm=TRUE)
mf = round(mf[1]-mf[2],2)
ar = tapply(petrels$R.Wing.Lth, petrels$Area, mean, na.rm=TRUE)
ar = round(ar[5]-ar[1],2)
```

- This doesn't mean that male birds have a `r -co["SexFemale"]`cm larger right wing length than females.

- In fact
    ```{r}
    tapply(petrels$R.Wing.Lth, petrels$Sex, mean, na.rm=TRUE)
    ```
    they're `r mf`cm larger across all areas.

## Petrels: Conclusion

- Similarly, birds in area 5 have `r co["Area5"]`cm larger right wing length than those in area 1 **after accounting for differences in sex**.

- Again,
    ```{r}
    tapply(petrels$R.Wing.Lth, petrels$Area, mean, na.rm=TRUE)
    ```
    they're `r ar`cm larger across both sexes.

## What about left wing length?

```{r, echo=FALSE, fig.align='center', fig.width=7, fig.height=5}
par(mar=c(4,4,0,1))
plot(R.Wing.Lth ~ L.Wing.Lth, data=petrels, ylab="Right wing length (cm)", xlab="Left wing length (cm)", col=col_points, pch=19)
```

## Adding in left wing length

```{r}
mod2 = lm(R.Wing.Lth ~ L.Wing.Lth + Sex + Area, data=petrels)
anova(mod2)
```

## Adding in left wing length{.smaller}

```{r, echo=FALSE}
summary(mod2)
co2 <- round(coef(mod2),2)
```

## Adding in left wing length

- From the ANOVA table, we see left wing length, sex and area are all still significant.

- From the summary table, however, we note that the direction of the effect of `Sex` has changed!

- It is now saying that females have `r co2["SexFemale"]`cm larger right wing lengths than males.

- Again, we've **adjusted for the other covariates**.

- **WEIRD!**

## Left and right wing lengths

```{r, echo=FALSE, fig.align='center', fig.width=7, fig.height=5}
col = c("#0000ff5F", "#FF00005F")

zones = matrix(c(2,0,1,3), ncol=2, byrow=T)
layout(zones, widths=c(6,1), heights=c(1,6))

par(mar=c(4,4,0.5,0.5))
plot(R.Wing.Lth ~ L.Wing.Lth, data=petrels, ylab="Right wing length (cm)", xlab="Left wing length (cm)", col=col[Sex], pch=19, asp=1)
legend("bottomright", legend=levels(petrels$Sex), fill=col, bty="n", cex=1.5)
lf=density(na.omit(petrels$L.Wing.Lth[petrels$Sex == "Female"]))
lm=density(na.omit(petrels$L.Wing.Lth[petrels$Sex == "Male"]))
rf=density(na.omit(petrels$R.Wing.Lth[petrels$Sex == "Female"]))
rm=density(na.omit(petrels$R.Wing.Lth[petrels$Sex == "Male"]))

par(mar=c(0,4,0,0.5))
plot(NULL, xlim=range(petrels$L.Wing.Lth, na.rm=TRUE), ylim=c(0,max(lf$y, lm$y)), axes=F, ann=FALSE)
lines(lm, col=col[1], lwd=2)
lines(lf, col=col[2], lwd=2)

par(mar=c(4,0,0.5,1))
plot(NULL, ylim=range(petrels$R.Wing.Lth, na.rm=TRUE), xlim=c(0,max(rf$y, rm$y)), axes=F, ann=FALSE)
lines(rm$y, rm$x, col=col[1], lwd=2)
lines(rf$y, rf$x, col=col[2], lwd=2)
```

## Left and right wing lengths

- In both left and right wing lengths, males are bigger.

- However, the difference is larger in the left wing lengths than the right.

- So after accounting for the left wing length (where we're incorporating that large difference in the sexes) we have to reduce the difference to get back down to the difference we see in the right wing lengths.

- The only way to do that is if the `Sex` coefficient is negative.

## Summary

- Always look at the `anova` table first, particularly if you have factors.

- Look at the `summary` table next, and interpret the P-values there for significant variables from the `anova` table.

- If variables are not significant, then you can remove them and re-run the `anova` and `summary` tables.

- **But this depends on what question we're answering!**

# Interactions

## Interactions

- Our current models have been assuming that each effect is **additive**.

- e.g. the difference in the right wing length between the sexes is common across all areas, and similarly the
area effects are common for males and females.

- This is often not the case.

## Example: Hypothetical drugs

- Suppose we have some medical condition and we have a numerical measure describing how good/bad a patient is.

- Suppose further that we have two drugs A and B to treat that condition, and that we have the option of combining them.

- Then there are 4 possible treatments: No drugs, just drug A, just drug B, or both drugs.

- Our current linear models, would assume that we can add the effect of drug A and B together for the "both drugs" case.

## Example: Hypothetical drugs

The model equation would be
$$
\mathsf{mean}(y) = \alpha + \beta_A z_A + \beta_B z_B
$$
where $z_A$ and $z_B$ are indicators for whether drug A or B are included, giving

Treatment     Mean
------------  ----
Neither drug  $\mathsf{mean}(y)=\alpha$
Just drug A   $\mathsf{mean}(y)=\alpha + \beta_A$
Just drug B   $\mathsf{mean}(y)=\alpha + \beta_B$
Both drugs    $\mathsf{mean}(y)=\alpha + \beta_A + \beta_B$

## Example: Hypothetical drugs

- For the combined drugs treatmet, we have $\mathsf{mean}(y)=\alpha + \beta_A + \beta_B$

- We're thus assuming that the individual benefit of each drug add together when combined.

- But, it might be that the combined effect of the drugs gives is smaller or larger than the sum of individual effects.

- We need a bit more flexibility in the model.

## Interactions: Hypothetical drugs

We add another indicator variable $z_{AB}$ which is 1 when both drugs are in use.
$$
\mathsf{mean}(y) = \alpha + \beta_A z_A + \beta_B z_B + \beta_{AB} z_{AB}
$$

The new indicator variable is just the product of the existing ones, as $z_{AB}=1$ only if both $z_A=1$ and $z_B=1$, and $z_{A} \times z_{B} = 1$ has the same property.

The means for the neither drug or single drug treatments are still the same, as $z_{AB}=0$ in those cases. The mean in the both drugs treatment is the only one that changes to
$$
\mathsf{mean}(y) = \alpha + \beta_A + \beta_B + \beta_{AB}.
$$
Thus, $\beta_{AB}$ can be used to measure whether or not the effects of the two drugs are additive or not.

## Example: Petrels

```{r, echo=FALSE, fig.width=9.5}
par(mar=c(5,5,1,0), cex.axis=1.5, cex.lab=1.5)
at = rep(1:2,6) + rep(1:6*2.5,each=2)
plot(R.Wing.Lth ~ interaction(Sex,Area), data=petrels, ylab="Right wing length (cm)", border=c("black", "red"), xaxt="n", xlab="",
     at=at)
mtext(rep(c("M","F"), 6),side=1, at=at, line=0.5)
mtext(paste("Area",1:6),side=1, at=1:6*2.5+1.5, line=2)
```
