---
title: "Lab 2: Petrels"
author: "CHANGE ME!"
output: html_document
---

## Introduction

In this notebook, we take another look at the data set on bycatch White-chinned Petrel *Procellaria aequinoctialis* caught predominantly by bottom longliners, tuna longliners, squid trawlers and general fish trawlers between 2000 and 2003.

We start by reading the data in, loading the `ggplot2` and `visreg` packages, and convert the `Area` variable to a factor.

```{r}
library(ggplot2)
library(visreg)
petrels = read.csv("http://www.massey.ac.nz/~jcmarsha/227215/data/petrels.csv")
petrels$Area = factor(petrels$Area)
```

In the following code chunk we remind ourselves what the distribution of `Sex` and `Area` are by producing a table, and also look at how the `R.Wing.Lth` changes with each of these variables using boxplots.

```{r}
```

## Linear modelling

Last semester we saw that there were differences in the size of birds between the areas - in particular, Area 5 had the largest birds. But, we also noticed that Area 5 also had the larger proportion of male birds, and male birds were larger than female. Thus, the size difference in Area 5 might just be down to having more male birds.

Today, we'll see if this is true, but modelling the data so that we can see the affect of area after accounting for the differences in sex.
