---
title: "Lab 2: Petrels"
author: "Jonathan Marshall"
output: html_document
---

## Introduction

In this notebook, we take another look at the data set on bycatch White-chinned Petrel *Procellaria aequinoctialis* caught predominantly by bottom longliners, tuna longliners, squid trawlers and general fish trawlers between 2000 and 2003.

We start by reading the data in, loading the `ggplot2` and `visreg` packages, and convert the `Area` variable to a factor.

```{r}
library(ggplot2)
library(visreg)
petrels = read.csv("http://www.massey.ac.nz/~jcmarsha/227215/data/petrels.csv")
petrels$Area = factor(petrels$Area)
```

In the following code chunk we remind ourselves what the distribution of `Sex` and `Area` are by producing a table, and also look at how the `R.Wing.Lth` changes with each of these variables using boxplots.

```{r}
table(petrels$Sex, petrels$Area)
```

We note that Area 5 has the highest proportion of male birds. Area 3 and 6 have the fewest birds.

```{r}
ggplot(na.omit(petrels)) +
  geom_boxplot(aes(x=Sex, y=R.Wing.Lth))
```

On average, the right wing length of male birds is a little bit higher than female birds, but otherwise the distributions are similar.

```{r}
ggplot(petrels) +
  geom_boxplot(aes(x=Area, y=R.Wing.Lth))
```

Area 5 birds are larger on average than the other areas, but a bit of overlap. Shapes and spreads are similar.

## Linear modelling

Last semester we saw that there were differences in the size of birds between the areas - in particular, Area 5 had the largest birds. But, we also noticed that Area 5 also had the larger proportion of male birds, and male birds were larger than female. Thus, the size difference in Area 5 might just be down to having more male birds.

Today, we'll see if this is true, but modelling the data so that we can see the affect of area after accounting for the differences in sex.

```{r}
lm1 <- lm(R.Wing.Lth ~ Area, data=petrels)
summary(lm1)
```

Our conclusion is that something in the model is important (overall P value is small). Area is the only thing in the model, so Area is important for explaining differences in right wing length (i.e. there's differences between areas). We can look at the coefficients to see where those differences are. e.g. Area 5 is largest, Area 6 is smallest.

```{r}
lm2 <- lm(R.Wing.Lth ~ Sex, data=petrels)
summary(lm2)
```

Overall P-value is small, so Sex is important. It explains less than 2% of the variation in right wing length. Males are larger than Females by about 3.1 mm.

```{r}
lm3 <- lm(R.Wing.Lth ~ Sex + Area, data=petrels)
summary(lm3)
```

After including Sex and Area in the model, the effect sizes of both have been moderated a little (suggesting that some of the effect of Area is explained by Sex, and some of the effect of Sex is explained by Area) but nonetheless the effect directions and magnitudes are similar: Males larger, Area 5 larger, Areas 2-4 and 6 are smaller compared to base line group.

Let's visualise the models:

```{r}
visreg(lm1)
```

```{r}
visreg(lm3)
```
The patterns we see for Area are the same in both models: Controlling for Sex has little effect on the overall conclusion. So, we can be confident that the Area 5 birds aren't larger on average just due to more male birds, as we've now controlled for that.

```{r}
plot(lm3)
```

No problem here: Linearity looks fine so does equal variance. Normality is good too. No observations with high influence, so all good!
