---
title: "How to weigh a donkey in Morocco"
author: "Jonathan Marshall"
output:
  pdf_document: default
  html_document: default
---

## Introduction

In this notebook we'll look at data collected on 385 donkeys used to transport goods to market in Morocco. It is a sample of the data collected for *Estimation of the liveweight and body condition of working donkeys in Morocco* by R.A. Pearson and M. Ouassat, published in the Veterinary Record, 1996. The following variables are available.

Variable   Description
--------   -----------
Sex        Sex (male/female)
Age        Age (years)
Bodywt     Weight (kg)
Heartgirth Girth at the heart (cm)
Umbgirth   Girth at the umbilicus (cm)
Length     Length from elbow to buttocks (cm)
Height     Height to the withers (cm)

We start by reading the data in from the web and looking at the first few rows:

```{r}
library(ggplot2)
donkey = read.csv("http://www.massey.ac.nz/~jcmarsha/227215/data/donkey.csv")
head(donkey)
```

We notice that the donkey data has a bunch of variables. Our interest is predicting body weight (`Bodywt`) using the other, more easily measureable things. Use of the `summary` command will give us some idea of how each of the columns are distributed:

```{r}
summary(donkey)
```

From this we see that the Bodyweight is somewhere between 52 and 222 kg, but that doesn't tell us much if we want to predict the body weight for a particular donkey. For that, we'll need to take advantage of the relationship between body weight and the other measures.

## Relationship between body weight and other measures

Let's see if there is a relationship between body weight and some of the other measurements. e.g. to assess body weight versus height we can use:

```{r}
ggplot(donkey, aes(x=Height, y=Bodywt)) + geom_point()
```

It's clear there is an increasing relationship between Bodyweight and Height as might be expected. This suggests we could use a particular donkey's height to get a better idea of its body weight.

```{r}
ggplot(donkey, aes(x=Length, y=Bodywt)) + geom_point()
```

```{r}
ggplot(donkey, aes(x=Umbgirth, y=Bodywt)) + geom_point()
```

```{r}
ggplot(donkey, aes(x=Heartgirth, y=Bodywt)) + geom_point(col='red')
```

```{r}
ggplot(donkey, aes(x=Sex, y=Bodywt)) + geom_boxplot()
```

From the above we can see that there is increasing relationships between all the measures and body weight, as might be expected. Heartgirth seems to have the strongest relationship, as the points are most clustered around the trend. The relationship is perhaps a little bit curved. The boxplot for sex suggests little difference between the bodyweights for the different sexes.

## Linear modelling of body weight

The below code block estimates a linear trend between body weight and heartgirth

```{r}
lm1 <- lm(Bodywt ~ Heartgirth, data=donkey)
summary(lm1)
```

From the output we can see that for each unit (cm) increase in Heartgirth, there is a corresponding 2.83kg increase in body weight on average. There's a little bit of uncertainty in this - a 95% confidence interval for the slope would be from $2.83 \pm 2\times 0.07 = (2.69, 2.97)$.

The equation that relates body weight to heartgirth would be
$$
\mathsf{mean(Bodyweight)} = -194.5 + 2.83 \times \mathsf{Heartgirth}
$$
The P-value for `Heartgirth` is testing whether the slope could be 0. It is very unlikely to be so, as the P-value is very close to zero. This makes sense from what we see in the graph - and also just from common sense - we'd expect the girth of the donkey influences it's body weight!

The $R^2$ output in the linear model summary is 0.8. This is a measure of the proportion of variation in body weight that is explained by our linear model (in this case, explained by heart girth). So it's saying that 80% of the variation we observe in donkey body weights is explained once we know the heartgirth of the donkey. So, once you know the heartgirth of the donkey, there's only about 20% of the variation in body weight left over.

Our best guess for what the body weight of a donkey with heartgirth 110cm would be: $$-194.5 + 2.83 \times 110 = 116.8$$kg. We can do this in RStudio using the `predict` function:

```{r}
# predict the weight of a donkey with heart girth 110cm
new_data <- data.frame(Heartgirth=110)
predict(lm1, new_data)
```

We can improve on this prediction by also incorporating the uncertainty that we have in the slope and intercept:

```{r}
predict(lm1, new_data, interval="confidence")
predict(lm1, new_data, interval="prediction")
```

The first prediction (with **confidence** interval) only accounts for the uncertainty in the mean body weight - recall the linear model equation is for the mean bodyweight across donkeys, not for an individual. Thus, you can consider this an estimate for the average bodyweight of donkeys with heartgirth 110cm. The second additionally incorporates the variation of individual donkeys around the average trend: You can consider it a **prediction** interval for an individual donkey with heartgirth 110cm. Obviously this one is wider.

## Model diagnostics

The residuals versus fitted plot is below.

```{r, eval=FALSE}
plot(lm1, which=1)
```

This shows that there is clear curvature in the residuals, so linearity may not hold. Also, we notice that the variation about the trend is increasing as we move from left to right (the points are fanning out more vertically) so equal variance also isn't very good. The implication is that our estimated relationship is likely a bit wrong.

We try a log-log transformation below to see if this helps (we could also try looking at a plot of Bodywt vs Heartgirth on a log-log scale, e.g. by adding `scale_x_log10()` and `scale_y_log10()` to our `ggplot`).

```{r}
lm2 <- lm(log(Bodywt) ~ log(Heartgirth), data=donkey)
summary(lm2)
```

The disadvantage with this model is it's less interpretable: The slope for heartgirth is now saying that a one unit change in the log of heartgirth produces a 2.58 unit change in the log of body weight. It isn't obvious what this means! If you rearrange the equation a bit by exponentiating each side we get something a little more interpretable:
$$
\begin{aligned}
\log(\mathsf{Bodywt}) &= -7.39 + 2.58 \times \log(\mathsf{Heartgirth})\\
\mathsf{Bodywt} &= 0.0006 \times \mathsf{Heartgirth}^{2.58}
\end{aligned}
$$
So this is suggesting there is a power relationship between bodyweight and heartgirth. This kinda makes sense - the heartgirth is measuring circumference, while weight would more likely be proportional to volume.

While we have the above disadvantages, perhaps we've nonetheless fixed the problem with the linear model assumptions not holding. Let's take a look:
```{r}
plot(lm2, which=1)
```

This looks much better - both linearity and equal variance look at though they hold. Also, if we look at the summary from `lm2` we note it has slightly better $R^2$ - it explains 82% of the variation instead of 80%. This, plus the improved model assumptions are probably enough to make up for the lack of interpretability.

When we do our predictions, we have to remember to exponentiate the result. The linear model will take care of doing the log of heartgirth for us, but won't 'unlog' the answer.
```{r}
new_data = data.frame(Heartgirth=110)
p1=predict(lm2, new_data, interval="confidence")
p2=predict(lm2, new_data, interval="prediction")
exp(p1)
exp(p2)
```
We note the predicted weight is a bit lower, and the uncertainty ranges are a bit tighter compared to the others. This makes sense, as the log/log model is explaining more of the variation, so there's less variation left over.

# Lab 2

## Visualising the model

Visualise the model fit using the `visreg` package

```{r}
library(visreg)
visreg(lm1)
```

We can see that the model fit (blue line) and uncertainty around it (grey band) is a pretty good representation
of the trend within the middle of the heartgirth range, but underestimates the weight of donkeys at either end of
the heartgirth range. This is because the trend is really curved, so the straight line fit doesn't really work.

The second model can be visualised in the same way
```{r}
visreg(lm2)
```

We can see that we have a curved relationship here, but the vertical axis is on the log scale rather than
the normal scale. Nonetheless, you can see that the points (which represent where the data are) are captured quite
well by the trend line, indicating we are modelling the trend quite well.

Use an exponential transformation to visualise the data and model on the natural scale:
```{r}
visreg(lm2, trans=exp, partial=TRUE, ylab="Body weight (kg)")
```

The above plot shows the data on the natural scale and better allows us to see the actual shape of the trend
which is slightly curved up. Notice the trend now captures (i.e. goes through) the values at the lower and upper
heartgirths now. We'd expect our predictions for animals with such heartgirths to be considerably better.

## Multiple regression
 
A linear model containing both heart and umbilical girths.
```{r}
lm3 <- lm(Bodywt ~ Heartgirth + Umbgirth, data=donkey)
summary(lm3)
```

The effect of heartgirth has changed as we're now also accounting for umbilical girth. Thus, each row represents the effect of the variable **after adjusting for other variables**. The P-value is thus testing: "Is this variable important to body weight after accounting for other variables?" This is different to "Is this variable important"!

Our conclusion would be that heartgirth and umbilical girth are both important for the bodyweight of the donkey after accounting for the other, as both P-values are significant.

Adding length and height to the model gives:
```{r}
lm4 <- lm(Bodywt ~ Heartgirth + Umbgirth + Length + Height, data=donkey)
summary(lm4)
```

After adjusting for heart girth, umbilical girth and length, there is little evidence that Height is important for body-weight as the P-value is quite large (the effect size of 0.25 kg/cm of Height could arise by chance about 6% of the time, so may not be genuine).

Note that this doesn't mean height is unimportant: If height is all we have, then we should use it:
```{r}
lm5 <- lm(Bodywt ~ Height, data=donkey)
summary(lm5)
```

We see Height is related to body weight. The difference is that in the previous model we'd accounted for a lot of the variation in bodyweight already by using heart and umbilical girths and length. Over and above those, the height didn't provide additional useful information. But absent of the other measures, height is ofcourse useful!

What about `Sex`? Recall that `Sex` wasn't important when we looked at the overall plots of weight vs sex last time. Replacing `Height` with `Sex` in `lm4` gives:

```{r}
lm6 <- lm(Bodywt ~ Heartgirth + Umbgirth + Length + Sex, data=donkey)
summary(lm6)
```

We see that Sex is related to body weight after accounting for the other variables, while Sex wasn't important in the boxplots we looked at, or in the model we looked at in lectures. Reason for this is once we have explained most of what contributes to weight (i.e. body size) the things that contribute a bit less, such as Sex can now be important for explaining the rest of the variation in weight.

## Visualising the model

Let's visualise this model:
```{r}
par(mfrow=c(2,2), mar=c(4,4,2,2))
visreg(lm6)
```

Here we can see that heartgirth, umbilical girth and length all result in increasing weights, and males are a bit heavier than females.

## Model diagnostics

And finally, assess model diagnostics

```{r}
par(mfrow=c(2,2), mar=c(4,4,2,2))
plot(lm6, add.smooth=FALSE)
```

We can see that linearity and equal variance don't seem to hold very well. Note that the `add.smooth=FALSE` here turns off the red curves which can be a bit better for judging the residual plots. Ideally we should include a transformation.
e.g. using a log transform for bodyweight is better:

```{r}
lm7 <- lm(log(Bodywt) ~ Heartgirth + Umbgirth + Length + Sex, data=donkey)
par(mfrow=c(2,2), mar=c(4,4,2,2))
plot(lm7, add.smooth=FALSE)
```

And this also explains about 1% more of the variation:

```{r}
summary(lm7)
```

The visreg plots look as though the model is fitting near the ends a bit better too

```{r}
par(mfrow=c(2,2), mar=c(4,4,2,2))
visreg(lm7, trans=exp, ylab="Bodyweight", partial=TRUE)
```
