---
title: 'Lab 2: Petrels'
author: "Jonathan"
output:
  pdf_document: default
  html_document: default
---

## Introduction

In this notebook, we take another look at the data set on bycatch White-chinned Petrel *Procellaria aequinoctialis* caught predominantly by bottom longliners, tuna longliners, squid trawlers and general fish trawlers between 2000 and 2003.

We start by reading the data in, loading the `ggplot2` and `visreg` packages, and convert the `Area` variable to a factor.

```{r}
library(ggplot2)
library(visreg)
petrels = read.csv("http://www.massey.ac.nz/~jcmarsha/227215/data/petrels.csv")
petrels$Area = factor(petrels$Area)
```

In the following code chunk we remind ourselves what the distribution of `Sex` and `Area` are by producing a table, and also look at how the `R.Wing.Lth` changes with each of these variables using boxplots.

```{r}
table(petrels$Sex, petrels$Area)
ggplot(subset(petrels, !is.na(Sex)), aes(x=Sex, y=R.Wing.Lth)) + geom_boxplot()
ggplot(petrels, aes(x=Area, y=R.Wing.Lth)) + geom_boxplot()
```

We notice that there is a higher proportion of males in area 5 compared to the other areas, and they have the largest right wing length. Males are also larger than females. Notice the use of `subset` to get rid of the NA values from `Sex`. This is to prevent a third boxplot consisting of the NA values (in addition to Male and Female).

## Linear modelling

Last semester we saw that there were differences in the size of birds between the areas - in particular, Area 5 had the largest birds. But, we also noticed that Area 5 also had the larger proportion of male birds, and male birds were larger than female. Thus, the size difference in Area 5 might just be down to having more male birds.

Today, we'll see if this is true, but modelling the data so that we can see the affect of area after accounting for the differences in sex.

A linear model for right wing length in terms of sex

```{r}
m1 <- lm(R.Wing.Lth ~ Sex, data=petrels)
summary(m1)
```

Our conclusion would be that males are about 3.1mm larger on average than females, which are 385.2mm on average, and that there is quite a bit of evidence for this (P=0.0006)

A model for right wing length in terms of area

```{r}
m2 <- lm(R.Wing.Lth ~ Area, data=petrels)
summary(m2)
```

Remember when assessing a single factor with more than 2 levels we can use the overall P-value. This is tiny, suggesting something in the model is important, so Area must be important to right wing length. We see that Area 5 is largest and area 6 is smallest.

And finally a model with both Sex and Area:

```{r}
m3 <- lm(R.Wing.Lth ~ Sex + Area, data=petrels)
summary(m3)
```

There definitely still seems to be an effect of area even after accounting for sex. Area 5 has largest and area 6 smallest wing lengths after accounting for sex. 

### Visualising the model

First we can look at the effect of Sex after accounting for area, where we see Males are larger.

```{r}
library(visreg)
visreg(m3, "Sex")
```

We can then look at the effect of area after accounting for Sex. It's useful to compare the results from the model with sex in it (m3) and the one with area alone (m2)

```{r}
par(mfrow=c(1,2), mar=c(4,4,2,2))
visreg(m3, "Area", main="Adjusted for Sex")
visreg(m2, main="Unadjusted")
```

By the looks there is no difference in the effect of Area between the two models, so accounting for Sex doesn't change the area effects. This suggests that sex is probably not explaining the area effect, so our concern about area 5 having more males is probably not an issue.

### Model diagnostics

```{r}
par(mfrow=c(2,2), mar=c(4,4,2,2))
plot(m3)
```

These look OK - no major evidence of non-linearity or non-constant variance, and normality looks OK (plus, large dataset!), so the model assumptions are likely fine.
