---
title:  227.215 Biostats, Lamb birthweights
author: CHANGE ME!
---

## Introduction

In this lab we'll look at a set of data on the birth weight of lambs, based on the data from a previous year's project, taken from a trial at Massey in the late 1990s. This will therefore be good practice for your project this year!

Mixed-aged, Border Leicester x Romney ($n=160$) ewes were randomly allocated after preganancy diagnosis to a factorial experiment involving two shearing treatments (shorn on day 70 vs not); two pregnancy ranks (single or twin); and four feeding regimes (HH, HM, MH, MM), where the first letter refers to feed during days 70-101 of pregancy, and the second during days 102-140.  The M (maintenance) feeding group were fed at a level calculated to maintain maternal conceptus-free weight, while the H (high) feeding group were fed to gain around 100g per day above this during the relevant period.

Of particular interest is how feed management and shearing during pregnancy influence lamb birth weight.  The data set consists of the following variables from 204 lambs.

Variable      Description
--------      -----------
`DamID`       Identifying number for the ewe.
`LambID`      Identifying number for the lamb.
`DateBirth`   Date of birth of the lamb.
`Rank`        Whether the lamb was Single or a Twin.
`Sex`         The sex of the lamb (Male, Female).
`BirthWeight` The weight of the lamb at birth (in kg).
`Feed`        The feed regime of the ewe (HH, HM, MH, HH).
`Shorn`       Whether the ewe was shorn during pregnancy (Yes, No).

We start by loading the libraries we need and reading in the data

```{r}
library(ggplot2)
library(visreg)
births <- read.csv("http://www.massey.ac.nz/~jcmarsha/227215/data/birthweight.csv")
head(births)
```

## Exploratory data analysis

Plot of birthweight versus rank

```{r}
ggplot(births, aes(x=Rank, y=BirthWeight)) + geom_boxplot()
```
