---
title: "227.215 Biostats, Lamb birthweights"
author: "Jonathan Marshall"
output: pdf_document
---

## Introduction

In this lab we'll look at a set of data on the birth weight of lambs, based on the data from a previous year's project, taken from a trial at Massey in the late 1990s. This will therefore be good practice for your project this year!

Mixed-aged, Border Leicester x Romney ($n=160$) ewes were randomly allocated after preganancy diagnosis to a factorial experiment involving two shearing treatments (shorn on day 70 vs not); two pregnancy ranks (single or twin); and four feeding regimes (HH, HM, MH, MM), where the first letter refers to feed during days 70-101 of pregancy, and the second during days 102-140.  The M (maintenance) feeding group were fed at a level calculated to maintain maternal conceptus-free weight, while the H (high) feeding group were fed to gain around 100g per day above this during the relevant period.

Of particular interest is how feed management and shearing during pregnancy influence lamb birth weight.  The data set consists of the following variables from 204 lambs.

Variable      Description
--------      -----------
`DamID`       Identifying number for the ewe.
`LambID`      Identifying number for the lamb.
`DateBirth`   Date of birth of the lamb.
`Rank`        Whether the lamb was Single or a Twin.
`Sex`         The sex of the lamb (Male, Female).
`BirthWeight` The weight of the lamb at birth (in kg).
`Feed`        The feed regime of the ewe (HH, HM, MH, HH).
`Shorn`       Whether the ewe was shorn during pregnancy (Yes, No).

We start by loading the libraries we need and reading in the data

```{r}
library(ggplot2)
library(visreg)
births <- read.csv("http://www.massey.ac.nz/~jcmarsha/227215/data/birthweight.csv")
head(births)
```

## Exploratory data analysis

Plot of birthweight versus rank

```{r}
ggplot(births, aes(x=Rank, y=BirthWeight)) + geom_boxplot()
```

This shows that rank has a clear effect on birthweight, with singles weight around 1kg more than twins.

Plots of birthweight versus sex, feed and shorn

```{r}
ggplot(births, aes(x=Sex, y=BirthWeight)) + geom_boxplot()
ggplot(births, aes(x=Feed, y=BirthWeight)) + geom_boxplot()
ggplot(births, aes(x=Shorn, y=BirthWeight)) + geom_boxplot()
```

There is a clear effect of sex, with males being heavier by about 400g or so, but less clear effects of feed and shorn.

It's possible that we can't see much difference in the feed treatment as there is too much within-treatment variation due to the other variables like rank and sex (the smaller lambs within each feed are likely twins, the largest likely singles).

Plotting subsets show us this might be right - the differences between feeds are more clear in the subsets.

```{r}
ggplot(subset(births, Rank == 'Twin'), aes(x=Feed, y=BirthWeight)) + geom_boxplot()
ggplot(subset(births, Rank == 'Single'), aes(x=Feed, y=BirthWeight)) + geom_boxplot()
ggplot(subset(births, Sex == 'Male'), aes(x=Feed, y=BirthWeight)) + geom_boxplot()
ggplot(subset(births, Sex == 'Female'), aes(x=Feed, y=BirthWeight)) + geom_boxplot()
```

Using facetting to produce small multiples is better:

```{r}
ggplot(births, aes(x=Feed, y=BirthWeight)) + geom_boxplot() + facet_wrap(~Rank)
ggplot(births, aes(x=Feed, y=BirthWeight)) + geom_boxplot() + facet_wrap(~Sex)
```

These are better as the plots are side-by side and the scales are the same, so we can see not only the differences between the feeds within each sex or rank, but also the differences between the ranks and sexes.

Using `facet_grid` instead is even better as it gives rows for the first variable and columns for the second.

```{r}
ggplot(births, aes(x=Feed, y=BirthWeight)) + geom_boxplot() + facet_grid(Sex ~ Rank)
```

Notice that the pattern of feed is similar in all panels: the `HM` treatment produces the lowest weights on average, while the other three are reasonably similar. Our conclusion is there is an effect of feed, but it's only really noticeable once we first account for rank and sex.

## Statistical modelling

A linear model assessing whether ewe nutrition is important for lamb birthweight. The linear model equation would be
$$
\mathsf{Birthweight} = \alpha + \beta_{HM} z_{HM} + \beta_{MH} z_{MH} + \beta_{MM} z_{MM}
$$
where $z_{HM}$ is an indicator variable with $z_{HM}=1$ when the ewe was in the $HM$ treatment group, and $z_{HM}=0$ otherwise. This gives $\mathsf{Birthweight}=\alpha$ when the ewe is in feed group HH, $\mathsf{Birthweight}=\alpha + \beta_{HM}$ when the ewe is in feed group HM and so on. Thus, $\beta_{HM}$ represents the difference in lamb birth weight between those ewes in feed group HH (the base-line) and those in feed-group HM.

```{r}
lm2 <- lm(BirthWeight ~ Feed, data=births)
anova(lm2)
summary(lm2)
```

The P-value of 0.11 from the ANOVA table suggests insufficient evidence for a difference in feed treatment on lamb birthweight. Interestingly, there does seem to be a significant difference between the HM feed group and the baseline (HH) in the summary table, but as the overall effect of feed is not significant, we shouldn't take this as being important (the summary table is assessing multiple hypotheses at once, and when you do that you have a higher chance of a false positive. In this case, the false positive is due to the HM and HH groups being the most distant possible pair we could look at: All other pairs of treatments are closer together than this one. It's not really a surprise that if we only look at the most distinct groups that there's evidence for a difference: We're being biased if we consider this is important while ignoring all the other pairs that show no difference!)

A linear model containing all variables is below.

```{r}
lm3 <- lm(BirthWeight ~ Sex + Rank + Feed + Shorn, data=births)
anova(lm3)
summary(lm3)
```

Based on the ANOVA table, we see all terms are significant, so all are important for birthweight.

The summary table suggests that
  - male lambs are on average 386g heavier than females.
  - twin lambs are just over 1 kg lighter than singles.
  - The HM feed treatment differs from the HH feed treatment, with ewes on the HM treatment being lighter by 450g than those on the HH treatment. The others seem similar to HH treatment.
  - Ewes that were shorn tend to have heavier lambs (by about 240g).

In earlier models both feed and shearing weren't significant. Now they are. The reason is that we've accounted for a bunch of the variation in the data by include Sex and Rank which we **know** affect birthweight. By accounting for some of the variation, we've reduced the remaining residual variation, thus being able to pick out the smaller between-group variation due to feed and shearing.

```{r}
new_data <- data.frame(Sex='Male', Rank='Twin', Shorn='No', Feed='HM')
predict(lm3, new_data, interval="prediction")
```

We can see a male twin born to an unshorn ewe on the HM feed treatment would be 4.45kg on average, and we'd expect to be between 2.84 and 6.06kg 95% of the time.

```{r}
new_data <- data.frame(Sex='Female', Rank='Single', Shorn='No', Feed='MM')
predict(lm3, new_data, interval="confidence")
```

We can see that female singles born to unshorn ewes on the MM feed treatment would be 5.56kg on avaerage, and we're 95% confident that the average is between 5.23 and 5.90kg.

We can visualise the model fit (essentially visualising what the summary table is telling us) with `visreg`

```{r}
library(visreg)
par(mfrow=c(2,2), mar=c(4,4,2,2))
visreg(lm3)
```

Notice this tells us the same thing as we got from the summary table: The HM group is lowest with the others somewhat similar to the HH group. Males are heavier, as are singles, and the shearing treatment seems to have a positive effect on weight. The blue lines are the model fit, and the grey bars are uncertainty (95\%).

## Interactions and model diagnostics

Adding an interaction between Shorn and Rank allows us to assess if the effect of shearing differs between those having twins and those having singles.

```{r}
lm4 <- lm(BirthWeight ~ Sex + Rank + Feed + Shorn + Rank:Shorn, data=births)
anova(lm4)
```

The `Rank:Shorn` variable is not significant from the anova table, so we'd conclude there isn't much evidence for the effect of shearing differing between singles and twins. We can see this in the visualisation as well

```{r}
visreg(lm4, "Shorn", by="Rank", gg=TRUE)
```

Notice the Yes vs No differences are about the same in each plot, especially when considering their uncertainties. This confirms what the anova table was telling us. You can also do an overlay plot for this which shows the same thing:

```{r}
visreg(lm4, "Shorn", by="Rank", overlay=TRUE, gg=TRUE)
```

Model diagnostics are shown below

```{r}
par(mfrow=c(2,2), mar=c(4,4,2,2))
plot(lm4)
```

We can see from the model diagnostics that
  - There is no trend in the residuals vs fitted plot suggesting
    the assumption of linearity is OK (residuals have mean 0 regardless of covariates).

  - The residuals seem to have constant variation in the residual vs fitted plot (or scale-location)
    suggesting the equal-variance assumption is satisifed.
  
  - The Normal-QQ plot suggests the residuals are mostly normal, only veering away at the lower tail.
  
  - There doesn't seem to be any residuals that have high influence (all points are inside the Cook's
    distance bands, as we can't even see the bands on the plot).
  
  - We can't assess the independence assumption with these plots. We'd be a little concerned that some of the
    residuals may not be independent (each pair of twins are likely similar). Ideally we'd take this into
    account in our model, but this is a bit beyond our current knowledge! Another way to 'fix' this, at the expense
    of lower power would be to remove one of the twins (at random) from the data-set and re-fit the model,
    but the problem is we'd be ignoring the within-ewe (between-twin) variation. Ignoring variation is usually
    a bad thing to do unless we know in advance that it's not important! In this case, between-twin variation
    would be important to know: If the twins are quite different, you'd want to know as maybe the feeding or
    shearing treatments don't apply equally to the twins (though I'm not sure how this would happen from a
    biological perspective!)
  
From this, we'd conclude the model assumptions are likely to hold, with the possible exception of independence. Our conclusions from the model are probably OK.
