---
title: "Linear models 2"
author: "Jonathan Marshall"
date: "18 June 2015"
output: 
  ioslides_presentation: 
    highlight: tango
    widescreen: true
    css: jm.css
---

```{r setup, echo=FALSE}
library(knitr)
opts_chunk$set(dev.args=list(bg='transparent'))
col_points <- "#7f577492"
col_dark   <- "#5f4354"
```
## Learning outcomes

- Adding more covariates.
- The Omnibus $F$-statistic.
- Models with factor variables.

# Example: Donkeys

## Relationships: Donkeys

```{r, echo=FALSE, fig.align="center", fig.width=8, fig.height=5}
donkey <- read.csv("http://www.massey.ac.nz/~jcmarsha/227215/data/donkey.csv")
par(mfrow=c(2,2), mar=c(4,4,1,1))
plot(Bodywt ~ Heartgirth, data=donkey, col=col_points, pch=19)
plot(Bodywt ~ Umbgirth, data=donkey, col=col_points, pch=19)
plot(Bodywt ~ Length, data=donkey, col=col_points, pch=19)
plot(Bodywt ~ Height, data=donkey, col=col_points, pch=19)
```

## Relationships: Donkeys

- There are reasonably strong increasing relationships between body weight and hearth girth, umbilical girth, length and height.

- It makes sense to use all of these variables to predict the body weight.

- That way we're likely to explain more of the variation in body weight.

- Predictions will be more precise (smaller confidence and prediction intervals).

# The Linear model|Multiple covariates

## Multiple covariates

- Adding other covariates is as simple as adding another term to the linear model equation.
$$
\mathsf{mean}(y) = \alpha + \beta_1 x_1 + \beta_2 x_2 + \cdots
$$
- We find least squares estimates for $\alpha$, $\beta_1, \beta_2, \ldots$ in the same way, by minimising the variance of residuals.
- As the model assumptions are defined in terms of the residuals, the complexity of the model doesn't really matter.
- We get additional lines in our output for each covariate we add to the model.

## Multiple covariates

We can formulate the linear model in terms of a **matrix equation**
$$
Y = X \beta + \epsilon
$$
where $Y$ is a vector of the dependent variable measurements, $X$ is a matrix of covariate values, $\beta$ is a vector of coefficients (to be estimated) and $\epsilon$ is a vector of residuals. The form $X$ takes is
$$
X = \left[\begin{array}{c c c c}
1 x_{11} \cdots x_{p1}\\
1 x_{12} \cdots x_{p2}\\
\vdots \vdots \ddots \vdots\\
1 x_{1n} \cdots x_{pn}
\end{array}\right]
$$
where $p$ is the number of covariates, and $n$ is the number of observations.

## Multiple covariates

- We don't want to just answer the question **"Which of the variables are important?"**

- Rather, we want to answer **"Which of the variables are important *after accounting for the other variables*?"**

- This is equivalent to: **"Which of the $\beta$'s are non-zero?"**

## Multiple covariates

```{r, comment="", echo=FALSE}
mod <- lm(Bodywt ~ Heartgirth + Umbgirth + Length + Height, data=donkey)
rs <- round(summary(mod)$r.squared,3)
fv <- round(summary(mod)$fstatistic[1],1)
summary(mod)
```

## Multiple covariates

- Heartgirth, umbilical girth and length are all significant after adjusting for other covariates.

- Height is not significant after accounting for the other covariates.

- The $R^2$ for the model is `r rs`, so the model is explaining `r 100*rs`% of the variation in body weight.

- What is the overall P-value telling us?

# The Omnibus F-test

## The Omnibus F-test

- One of the first things we want to know is whether anything at all in our model is useful.

- The hypothesis to test is "Are any of the $\beta$'s non-zero?"

- We can do this by comparing the variation in the residuals to the total variation in $y$.

- We use the F-statistic for this.
$$
F = \frac{(SS_\mathsf{total} - SS_\mathsf{res})/p}{SS_\mathsf{res}/(n-p-1)}
$$
where $p$ is the number of covariates, and $n$ the number of observations.

## The Omnibus F-test
$$
F = \frac{(SS_\mathsf{total} - SS_\mathsf{res})/p}{SS_\mathsf{res}/(n-p-1)}
$$

- On the numerator we have a measure of variance explained by the model.

- On the demoninator we have a measure of variance left over in the residuals.

- The ratio of these, $F$ will increase as we explain more variation in the model.

- Under the null hypothesis, the model doesn't explain anything, so we'd expect $F=0$.

- We can use the $F$&nbsp;distribution to figure out how likely the $F$ from our sample
would arise by chance, giving us the P-value.

## The Omnibus F-test

```
Coefficients:
              Estimate Std. Error t value Pr(>|t|)    
(Intercept) -222.33956    8.01411 -27.744  < 2e-16 ***
Heartgirth     1.76223    0.12784  13.784  < 2e-16 ***
Umbgirth       0.37418    0.06726   5.563 5.01e-08 ***
Length         0.87580    0.11531   7.595 2.41e-13 ***
Height         0.25407    0.13749   1.848   0.0654 .  
---
Signif. codes:  0 ‘***’ 0.001 ‘**’ 0.01 ‘*’ 0.05 ‘.’ 0.1 ‘ ’ 1

Residual standard error: 9.476 on 380 degrees of freedom
Multiple R-squared:  0.8519,	Adjusted R-squared:  0.8503 
### <b>
F-statistic: 546.4 on 4 and 380 DF,  p-value: < 2.2e-16
### </b>
```

An F value of `r fv` is really unlikely to arise by chance, so our model is telling us something about body weight.

## The F distribution

```{r, echo=FALSE}
library(shiny)
shinyAppDir("../shiny/fstatistic")
```

# Linear models with factors

## Linear models with factors

- The linear model equation is describing the **mean** of the outcome variable given the value of the covariates.

- So we can use a linear model to estimate a mean just by not including any covariates, e.g. by using the model equation
$$
\mathsf{mean}(y) = \alpha
$$

- When we have a factor, or categorical covariate, the linear model will estimate the mean within each group.

## Factors with two levels.

- For a factor with two levels, we can do this by introducing an **indicator variable** $z$ as a covariate, where
$$
z_i = \left\{
\begin{array}{ll}
0 & \textrm{if }x_i\textrm{ is in level 1}\\
1 & \textrm{if }x_i\textrm{ is in level 2}
\end{array}\right.
$$
- The equation is then
$$
\mathsf{mean}(y) = \alpha + \beta z
$$

## Factors with two levels.

- The equation is then
$$
\mathsf{mean}(y) = \alpha + \beta z
$$
- For observations in level 1, $z=0$ so the equation is just
$$
\mathsf{mean}(y) = \alpha
$$
- For observations in level 2, $z=1$, so the equation is
$$
\mathsf{mean}(y) = \alpha + \beta
$$
- Thus, $\beta$ represents the **difference** between level 1 and 2.

- Testing for a difference between groups is then just testing whether $\beta = 0$.

## Example: Donkeys {.smaller}

```{r}
mod <- lm(Bodywt ~ Sex, data=donkey)
summary(mod)
```

## Example: Donkeys

```{r}
t.test(Bodywt ~ Sex, data=donkey, var.equal=TRUE)
```

<div class="centered">
**Linear model gives same results!**
</div>

## What about three levels?

- For two levels we introduced an indicator variable that was 1 when we were in level 2 and 0 otherwise.

- For three levels we introduce two indicator variables:

    - $z_2 = 1$ if the observation is in level 2, $z_2 = 0$ otherwise.

    - $z_3 = 1$ if the observation is in level 3, $z_3 = 0$ otherwise.

- The linear model is then
$$
\mathsf{mean}(y) = \alpha + \beta_2 z_{2} + \beta_3 z_{3}
$$
where $\beta_2$ is the difference between level 1 and level 2, and $\beta_3$ is the difference between level 1 and level 3.

## What about three levels?

- The linear model is then
$$
\mathsf{mean}(y) = \alpha + \beta_2 z_{2} + \beta_3 z_{3}
$$

- The primary hypothesis to test is "Is there any difference between the three levels?".

- This is the same as "Are any of the $\beta$'s zero".

- The Omnibus F-test!

## Example: Petrels

- First step, a plot.

- Second step, what the model is estimating.

- Third step, what the F-statistic represents.

- Fourth step, perhaps an app for that?

## The `anova` function

